(Re-)bridging the anatomy ontologies with SSSOM

### 2024-06-06 presentation.tex
Shorten the section on the previous system.
The presentation should fit in 15 minutes, so we can't afford to spend 3
slides explaining how the previous integration system was working.

  This talk is about how this integration process works.}

@@

\section{Current system}

\begin{frame}
  \frametitle{Maintaining mappings as cross-references}

  \begin{block}{}\ttfamily
[Term]

@@

def: "The caudal, pulsatile region of the dorsal vessel of the arthropod circulatory system."

@@

  Limitations:
  \begin{itemize}
    \item Overloading of the \emph{oboInOwl:hasDbXref} annotation
    \item No metadata
    \item Mappings are embedded within the ontology and must be extracted
  \end{itemize}

  \note{In the current system, mappings are represented and maintained as
  \emph{cross-references}, that is, as a \emph{oboInOwl:hasDbXref} annotation
  that is carried by the taxon-neutral term in Uberon and that points to the
  taxon-specific term in another ontology -- as shown here, where Uberon's
  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
  FBbt's term for the fly heart that we have seen before.

  That's a pretty simple way of representing mappings, but it has several
  drawbacks.

  First, the \emph{oboInOwl:hasDbXref} annotation has been used for a lot of
  things since its introduction, and no longer has a clear and unambiguous
  meaning -- what it means depends on where it is used.

  Second, all this representation allows us to say is that there is a mapping,
  but we can't know anything else about said mapping. Why does that mapping
  exist? Who asserted it? When? On what basis? This is simply not recorded.

  Third, the mappings are not available separately from the entire ontology.
  This makes reusing those mappings more difficult than it should be
  (difficult to Find, Access and Reuse).}

\begin{frame}
  \frametitle{From cross-references to bridge files}

  \begin{columns}
    \begin{column}{.3\textwidth}
      \begin{tikzpicture}
        \node [File] (uberon) at (0,0)     {uberon.obo};
        \node [File] (script) at (0,-1)    {make-bridge-ontology-from-xrefs.pl};
        \node [File] (bridge) at (0,-2)    {uberon-bridge-to-fbbt.obo};
        \node [File]          at (0, -2.5) {uberon-bridge-to-wbbt.obo};
        \node [File]          at (0, -3)   {uberon-bridge-to-ma.obo};
        \node [File]          at (0, -3.5) {uberon-bridge-to-....obo}; 
        \draw[->] (uberon) -- (script) (script) -- (bridge);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.68\textwidth}
      \begin{block}{uberon-bridge-to-fbbt.obo}\ttfamily
[Term]

@@

intersection\_of: part\_of NCBITaxon:7227
      \end{block}

      Can be used to integrate FBbt into Uberon
    \end{column}
  \end{columns}

  \medskip
  Limitations:
  \begin{itemize}
    \item Highly Uberon-specific code -- cannot be reused outside of Uberon
    \item Dependent on the OBO format
  \end{itemize}

  \note{Whenever we make a new Uberon release, the mappings (stored as
  cross-references within Uberon) are extracted by a specialised Perl script
  to produce a set of \emph{bridge files} (one for each species-specific
  ontology). The bridge files are a OBO materialisation of the mappings; they
  contain minimalist frames that serve to rattach a species-specific term to a
  core Uberon term, as in this example, which states that FBbt:00003154 (the
  fly heart) is equivalent to a Uberon's ``dorsal vessel heart'' that is part
  of the \emph{Drosophila melanogaster} taxon.

  The main problem with this derivation is that the aforementioned script is
  highly specific to Uberon. Most of its behaviours are hard-coded, and the
  script is virtually impossible to re-use outside of Uberon. This means,
  among other things, that third-party projects that might need those bridges,
  for one reason or another, are entirely dependent on Uberon to produce them.
  One example of such a project is Ubergraph.

  Another limitation is that the script only works with the OBO format, both
  for its input and its output.}

@@

\begin{frame}
  \frametitle{Integration pipeline}

  \begin{columns}
    \begin{column}{.44\textwidth}
      \begin{tikzpicture}[node distance=.5cm]
        \node [Entity] (uberon) {Uberon};
        \node [Process,below right=of uberon] (script) {Custom script};
        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
                                                                ontologies};
        \node [Entity,below=of script] (cob) {Bridge files};
        \node [Process,below=of cob] (merge) {OWL merge};
        \node [Entity,below=of merge] (cm) {Multi-species ontology};
        \draw [->] (uberon.east) -| (script.north) (script.south) -- (cob.north);
        \draw [->] (cob) -- (merge) (merge) -- (cm);
        \draw [->] (uberon.south) |- (merge.west);
        \draw [->] (ssao.south) |- (merge.east);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.55\textwidth}
      Limitations:
      \begin{itemize}
        \item Lack of flexibility
          \begin{itemize}
            \item Cannot integrate foreign mappings
            \item Cannot generate a single bridge file
          \end{itemize}
        \item Mappings not available independently
        \item Not generalizable to other ontologies
      \end{itemize}
    \end{column}
  \end{columns}

  \note{Finally, here is an outline of the entire integration process, what
  happens when we do a Uberon release. From the main Uberon source file, we
  create the bridge files using the aforementioned script, and then we merge
  together: (1) Uberon itself, (2) the bridge files, and (3) the
  species-specific ontologies, to produce the integrated, multi-species
  ontology.

  While the process looks superficially simple (though it is, in fact, quite
  complex, it's just that most of the complexity is hidden behind that
  script), it is also severely limited.

  It is inflexible, in that the only possible source of mappings to generate
  the bridge files are the xrefs in Uberon itself -- we cannot integrate
  mappings coming from another source (which means we cannot delegate the
  maintenance of the mappings to the species-specific ontologies). Also, we
  cannot generate a single bridge file on-demand -- the script can only be
  used to generate all the bridge files in one step. It does not allow to
  publish the mappings themselves.

  It cannot be applied to other ontologies. There might be other projects that
  need to integrate several ontologies together, using mappings to connect
  terms across ontologies, but those projects cannot benefit from the work
  done in Uberon.}

\section{New approach}

\begin{frame}
  \frametitle{Principles}

  \begin{enumerate}
    \item Clear separation between
      \begin{itemize}
        \item cross-ontology \emph{mappings} (COM)
        \item cross-ontology \emph{bridges} (COB)
      \end{itemize}
    \item Represent COM using a standard format intended for that purpose
      \begin{itemize}
        \item allowing for mapping-specific metadata
        \item allowing for meaningful mapping predicates
      \end{itemize}
    \item Make it easy to derive COB from COM
      \begin{itemize}
        \item without requiring Uberon-specific tooling
      \end{itemize}
  \end{enumerate}

  \note{Here are the key principles that dictated the design of the new
  approach.

  First, we want a clear separation between the \emph{mappings} and the
  \emph{bridges} that are derived from the mappings. In our current system,
  the mappings only exist as annotations within the ontology, and are not
  really treated as ``first-class objects''. In our new system, we want to
  focus on the mappings rather than the bridges, because the mappings are the
  actual informations that we are maintaining -- the bridges are merely a
  derivatiom from them. In particular, we want to be able to publish the
  mappings themselves, independently of the bridges.

  Second, we want to store the mappings using a format that is both standard
  and specifically intended for representing mappings. That format should,
  among other things, make it possible to store mapping-specific metadata (who
  asserted the mapping, etc.), make it possible to give precise meaning to a
  mapping (not just saying that two terms are mapped, but what is the nature
  of the link between them).

@@

@@

  \note{This leads us to our now, SSSOM-based integration pipeline for Uberon,
  which now has three distinct inputs: (1) Uberon itself, (2) the mappings
  between Uberon and the taxon-specific ontologies, and (3) the taxon-specific
  ontologies.

>>>>

  This talk is about how this integration process works, and how we recently
  overhauled it to make it easier to maintain, more robust, and FAIRer.}

@@

\section{Limits of the previous system}

\begin{frame}
  \frametitle{Overview of the previous integration system}

  \begin{columns}
    \begin{column}{.44\textwidth}
      \begin{tikzpicture}[node distance=.5cm]
        \node [Entity] (uberon) {Uberon};
        \node [Process,below right=of uberon] (script) {Custom script};
        \node [Entity,above right=of script,align=left] (ssao) {Taxon-specific\\
                                                                ontologies};
        \node [Entity,below=of script] (cob) {Bridge files};
        \node [Process,below=of cob] (merge) {OWL merge};
        \node [Entity,below=of merge] (cm) {Multi-species ontology};
        \draw [->] (uberon.east) -| (script.north) (script.south) -- (cob.north);
        \draw [->] (cob) -- (merge) (merge) -- (cm);
        \draw [->] (uberon.south) |- (merge.west);
        \draw [->] (ssao.south) |- (merge.east);
      \end{tikzpicture}
    \end{column}
    \begin{column}{.53\textwidth}
      \begin{block}{Mappings maintained as xrefs}\scriptsize\ttfamily
[Term]

@@

@@

      \begin{block}{Xrefs transformed to bridge files}\scriptsize\ttfamily
[Term]

@@

intersection\_of: part\_of NCBITaxon:7227
      \end{block}
    \end{column}
  \end{columns}

  \begin{itemize}
    \item Mappings not available to third parties
    \item Pipeline highly Uberon-specific, inflexible and not reusable
  \end{itemize}

  \note{Mappings are represented and maintained as \emph{cross-references}
  that are carried by the taxon-neutral terms in Uberon and that point to the
  taxon-specific terms in other ontologies -- as shown here, where Uberon's
  ``dorsal vessel heart'' term carries a ``xref'' to FBbt:00003154, which is
  FBbt's term for the fly heart.

  When building Uberon, those cross-references are extracted by a custom
  script that turns them into \emph{bridge files}, which contains OWL
  materialisations of the mappings, in the form of equivalence axioms (for
  example here, an axiom stating the FBbt's ``heart'' is equivalent to an
  Uberon's ``dorsal vessel heart'' that is part of the \emph{Drosophila
  melanogaster} taxon).

  This system has severe limitations, one of the most important being that the
  mappings are directly embedded within the ontology itself and are not
  independently available. In FAIR terms, this makes the mappings much more
  difficult to find, access and reuse than we would like.

  Furthermore, the custom script that generates the bridge files from the
  cross-references is highly specific to Uberon; its behaviour are completely
  hardcoded to fit Uberon's use case, so the integration logic cannot possibly
  be reused by other projects.}

@@

\begin{frame}
  \frametitle{Goals for a new approach}

  \begin{enumerate}
    \item Cross-ontology mappings (COM) as first-class entities
      \begin{itemize}
        \item available independently of the ontology itself
      \end{itemize}
    \item Represent mappings using a standard format intended for that purpose
      \begin{itemize}
        \item allowing for mapping-specific metadata
        \item allowing for meaningful mapping predicates
      \end{itemize}
    \item Make it easy to derive OWL bridge files from mappings
      \begin{itemize}
        \item without requiring Uberon-specific tooling
      \end{itemize}
  \end{enumerate}

  \note{In response to the limits of the previous system, here are the key
  principles that dictated the design of the new approach.

  First, we want to treat mappings as ``first-class entities'', on the same
  level as the terms of the ontology. Among other things, this means that they
  must be publicly available on their own, without requiring people interested
  in them to forcefully extract them from the ontology.

  Second, we want to store the mappings using a format that is both standard
  and specifically intended for representing mappings. That format should,
  among other things, make it possible to store mapping-specific metadata, and
  to give precise meaning to a mapping (not just saying that two terms are
  mapped, but what is the nature of the link between them).

@@

\section{Implementation}

@@

  \note{This leads us to our new, SSSOM-based integration pipeline for Uberon,
  which now has three distinct inputs: (1) Uberon itself, (2) the mappings
  between Uberon and the taxon-specific ontologies, and (3) the taxon-specific
  ontologies.
